Understanding meta-analysis through data simulation with applications to power analysis

### 2023-09-06 documents/paper/paper.Rmd
adding section on tutorial usage

keywords          : "meta-analysis, monte carlo simulations, power analysis"
wordcount         : "7782"
bibliography      : "../files/references.bib"
floatsintext      : yes
figsintext        : yes
linenumbers       : "`r params$linenumbers`"
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
numbersections    : false
nocite: |
  @r-lang
header-includes:
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1,footnotesize}}
  - \usepackage{float}
  - \usepackage{orcidlink}
  - \newcommand{\rev}[1]{\textcolor{red}{#1}}
documentclass     : "apa6"
output            :
  papaja::apa6_pdf:
    latex_engine: xelatex
params: 
  linenumbers: "yes"
  finalsection: TRUE
---

@@

@@

sim_grid |> 
  ggplot(aes(x = k, y = power)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "firebrick4") +
  geom_line() +
  geom_point(size = 4, shape = 21, color = "white", fill = "black", stroke = 3) +
  theme_paper() +
  ylim(c(0, 1)) +
  xlab("Number of studies (k)") +
  ylab("Power") +
  ggtitle(title)
```

@@

(ref:img-equal-vs-random)  The difference between the assumptions of the *equal* and *random-effects model*. Each distribution depicts the sampling distribution of $k = 4$ hypothetical studies ($i = 1$, $2$, ..., $4$) with a certain observed effect size $y_{i}$ (red squares), sampling variability $\sigma_{\epsilon_i}^{2}$ and the 95% confidence interval (black segment). The *equal-effects* plot on the left suggests that each observed effect size has the same underlying true effect $\theta$ (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be close to the real effect ($\theta$). The *random-effects model* on the right suggests that beyond the error term ($\epsilon_{i}$), each real effect size is composed of a fixed part (now $\mu_{\theta}$) and a random part ($\delta_{i}$) sampled from a normal distribution with mean zero and variance $\tau^{2}$. When $\tau^{2}$ is zero, the random-effects reduces to a equal-effects model.

@@

(ref:img-metaregression-bin) Graphical representation of a *random-effects meta-regression* model with a categorical predictor (Condition A and Condition B). Each grey distribution represents the sampling distribution of included studies. The dotted line is the average effect (i.e., random-effects model without moderators). The effect size differs between conditions A and B, including the *condition* moderator explaining part of the total heterogeneity (red plus green segments). The green segments depict the explained heterogeneity, and the red segments the residual (unexplained) heterogeneity.  The red squares are simulated observed effect sizes from the sampling distributions.

@@

(ref:plot-power-analysis) Results from the *random-effects model* power analysis. The x-axis depicts the number of studies ($k$), and the y-axis the estimated power. The red dotted line is the 80% power level, usually considered a good value for power analysis.

>>>>

keywords          : "meta-analysis, monte carlo simulations, power analysis"
wordcount         : "7866"
bibliography      : "../files/references.bib"
floatsintext      : yes
figsintext        : yes
linenumbers       : "`r params$linenumbers`"
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
numbersections    : false
nocite: |
  @r-lang
header-includes:
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1,footnotesize}}
  - \usepackage{float}
  - \usepackage{orcidlink}
  - \newcommand{\rev}[1]{\textcolor{red}{#1}}
documentclass     : "apa6"
output            :
  papaja::apa6_pdf:
    latex_engine: xelatex
params: 
  linenumbers: "yes"
  finalsection: TRUE
---

@@

The aim of the tutorial is not providing a complete theoretical introduction to meta-analysis but presenting the core topics using a simulation-based approach. Readers expert in conducting meta-analyses could benefit more from the proposed approach in terms of simulation setup and coding strategies. Readers without experience in meta-analysis could benefit both from the theoretical introduction and the simulation approach. However, for a better and extensive overview of meta-analysis topics the reader could refer to meta-analysis textbooks [e.g., @Harrer2019-rl; @Borenstein2009-mo].

@@

sim_grid |> 
  ggplot(aes(x = k, y = power)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "#CC79A7") +
  geom_line() +
  geom_point(size = 4, shape = 21, color = "white", fill = "black", stroke = 3) +
  theme_paper() +
  ylim(c(0, 1)) +
  xlab("Number of studies (k)") +
  ylab("Power") +
  ggtitle(title)
```

@@

(ref:img-equal-vs-random)  The difference between the assumptions of the *equal* and *random-effects model*. Each distribution depicts the sampling distribution of $k = 4$ hypothetical studies ($i = 1$, $2$, ..., $4$) with a certain observed effect size $y_{i}$ (pink squares), sampling variability $\sigma_{\epsilon_i}^{2}$ and the 95% confidence interval (black segment). The *equal-effects* plot on the left suggests that each observed effect size has the same underlying true effect $\theta$ (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be close to the real effect ($\theta$). The *random-effects model* on the right suggests that beyond the error term ($\epsilon_{i}$), each real effect size is composed of a fixed part (now $\mu_{\theta}$) and a random part ($\delta_{i}$) sampled from a normal distribution with mean zero and variance $\tau^{2}$. When $\tau^{2}$ is zero, the random-effects reduces to a equal-effects model.

@@

(ref:img-metaregression-bin) Graphical representation of a *random-effects meta-regression* model with a categorical predictor (Condition A and Condition B). Each grey distribution represents the sampling distribution of included studies. The dotted line is the average effect (i.e., random-effects model without moderators). The effect size differs between conditions A and B, including the *condition* moderator explaining part of the total heterogeneity (pink plus green segments). The green segments depict the explained heterogeneity, and the pink segments the residual (unexplained) heterogeneity. The pink squares are simulated observed effect sizes from the sampling distributions.

@@

(ref:plot-power-analysis) Results from the *random-effects model* power analysis. The x-axis depicts the number of studies ($k$), and the y-axis the estimated power. The pink dotted line is the 80% power level, usually considered a good value for power analysis.
